---
fontsize: 9pt
format:
  typst:
    papersize: a4
    margin:
      x: 1.2cm
      y: 1.2cm
---

# Data mining and statistical tests

## A. Clinical Problem

**RQ:** Is proximal arm non-use (**PANU**) related to shoulder non-use (**SANU**) and or elbow non-use (**EENU**) after stroke?

### Variables

After a stroke, a patient may compensate for reduced shoulder or elbow use by leaning the trunk forward more. Therefore, we want to know whether:

- more **shoulder non-use** is associated with more **proximal arm non-use**
- more **elbow non-use** is associated with more **proximal arm non-use**

This can be studied using the **correlations between PANU, SANU and EENU**.

| Variable | Description | Unit | Range |
|----------|---------------------------------------|---------|------------------|
| SANU | shoulder flexion the patient can use with the trunk restrained but does not use spontaneously. | \% | 0-100 |
| EENU | elbow extension the patient can use with the trunk restrained but does not use spontaneously. | \% | 0-100 |
| PANU | shoulder and elbow movement together that the patient can use with the trunk restrained but does not use spontaneously. | \% | 0-100 |

**+** = non-use : patient used more movement in the restrained condition than in the spontaneous condition

**-** = over-use : patient used more movement spontaneously than in the restrained condition

**0** → no difference between the two conditions.

PANU (%) = (hand displacement − trunk displacement) / hand displacement

## B. Experimental Methodology

Participants with a previous stroke performed a **seated reaching task with the affected arm** while their movement was recorded using motion capture.

### Condition 1: Spontaneous

- The participant reaches for the cone using their natural movement strategy.
- Trunk compensation is allowed.
- This shows **how the participant normally chooses to perform the movement**.

### Condition 2: Trunk Restrained

- The participant performs the same reach while keeping the trunk still.
- This reduces forward trunk compensation.
- The participant must rely more on **shoulder and elbow movement**.
- This shows **what the affected arm can do when trunk compensation is restricted**.

## C. Data: How Non-Use Is Identified

- If the participant produces more shoulder and or elbow movement when the trunk is restrained than during spontaneous reaching this shows that they were **capable of using that movement but did not use it spontaneously**.
- **Non-use =** movement capacity available − movement used spontaneously

| Variable | Correlation Prediciton | Why? | Unit | Range |
|---------|---------------|-------------------------------------|-------|---------|
| SANU | Positive | More shoulder non-use should mean more overall proximal-arm non-use. | \% | 0-100 |
| EENU | Positive | More elbow non-use should mean more overall proximal-arm non-use. | \% | 0-100 |
| PANU | Positive | Both contribute to reaching, but one joint can partly compensate for the other. | \% | 0-100 |

## 1. Import data

```{r}
# 1. Import libraries
library(tidyverse) # tools for plotting and data analysis
```

```{r # Import libraries and import the data}

# 2. Import the dataset
nonuse <- read_csv("data/NonUse.csv")

# View the first 6 rows to make sure it was imported correctly 
head(nonuse)
```

```{r}
# summary of the data
summary(nonuse)
```

## 2. Check the imported data

```{r}
str(nonuse) # structure of data
```

## 3. Clean data

The data set contains missing values, only complete pairs of observations were used for each analysis. Thus the data must be cleaned before performing any correlation analysis.

```{r}

# Keep only complete observations for all three variables
nonuse_clean <- nonuse %>%
  select(PANU, SANU, EENU) %>%
  drop_na() # removes rows with NA
```

```{r}
head(nonuse_clean) # check the new data table
```

## 4. Distribution of data

Before the correlation analysis, boxplots were used to inspect the distribution, spread, and possible outliers of PANU, SANU, and EENU.

```{r}

# 1. Convert the data from wide to long format
nonuse_long <- nonuse_clean %>%
  pivot_longer(cols = c(PANU, SANU, EENU), names_to = "Variable", values_to = "Value")
  # convert the data into a two columns one with variable names and the other with values because ggplot requires this data formating 

# 2. Create one boxplot for all three variables
ggplot(nonuse_long, aes(x = Variable, y = Value)) + 
  geom_boxplot() +
  labs(title = "Distribution of PANU, SANU and EENU", x = "Variable", y = "Value") +
  theme_minimal()

```

## 5. Correlation analysis

- Visualize `PANU–SANU` and `PANU–EENU`

- Calculate the correlations

- Interpret the correlation coefficients

- Perform the two linear regressions

- Extract the regression equations

- Write a scientific-results sentence

- Discuss limitations

### a). Visualize relationships: scatterplot of PANU-SANU and PANU-EENU

Q: As SANU or EENU changes, does PANU tend to change as well?

- `SANU or EENU` as the **x-variable**

- `PANU` as the **y-variable**

```{r}
# 1. Create a scatterplot for PANU-SANU

ggplot(nonuse_clean, aes(x = SANU, y = PANU)) +
  geom_point() + # represent each observation as a point
  labs(title = "Relationship between PANU and SANU", x = "SANU", y = "PANU") +
  theme_minimal()
```

```{r}
# 2. Create a scatterplot for PANU-EENU

ggplot(nonuse_clean, aes(x = EENU, y = PANU)) +
  geom_point() + 
  labs(title = "Relationship between PANU and EENU", x = "EENU", y = "PANU") +
  theme_minimal()
```

### Interpretation

> The PANU–EENU relationship appears more clearly positive, while the PANU–SANU relationship shows greater dispersion. Several observations are far from the main cluster, which may influence the correlation. The strength of these relationships will be quantified using correlation coefficients.

### b). Linear regression: PANU-SANU and PANU-EENU

General form: `y = ax + b`

***Metrics:***

1.  $p$ \< 0.05 $\rightarrow$ statistically significant
2.  $R^2$ = proportion of variation explained by the linear model
    - $R^2$ $\rightarrow$ (0 to 1)
3.  $r$ = direction + strength of linear association
    - $r$ $\rightarrow$ (-1 to +1)

```{r}
# 1. Linear regression: PANU-SANU

model_SANU <- lm(PANU ~ SANU, data = nonuse_clean) # store the linear model lm(). 
# PANU = left side of equation (variable we want to predict), SANU = right side of equation (predictor variable)

summary(model_SANU) # show the main regression results (intercept b, slope a)

# 2. Linear regression: PANU-EENU

model_SANU <- lm(PANU ~ EENU, data = nonuse_clean)

summary(model_SANU) 
```

<table>
<thead>
<tr>
<th><p>PANU-SANU</p></th>
<th><p>PANU-EENU</p></th>
</tr>
</thead>
<tbody>
<tr>
<td><p>PANU = a * SANU + b </p></td>
<td><p>PANU = a * EENU + b </p></td>
</tr>
<tr>
<td><p><code>PANU = 0.10093 * SANU + 8.86819</code></p>
<ul>
<li><p>Slope (a) = 0.10093 <span class="math inline">→</span>  (a &gt;0) positive relationship </p></li>
<li><p>Intercept (b) = 8.86819</p></li>
</ul></td>
<td><p><code>PANU = 0.39851 * EENU + 5.67295</code></p>
<ul>
<li><p>Slope (a) = 0.39851 <span class="math inline">→</span>  (a &gt;0) positive relationship </p></li>
<li><p>Intercept (b) = 5.67295</p></li>
</ul></td>
</tr>
<tr>
<td><p><code>p = 0.171</code> </p>
<ul>
<li><p>0.171 &gt; 0.05 <span class="math inline">→</span> slope is not statistically significant. Although the slop is positive, the data do not provide sufficient evidence to determine a linear relationship between PANU and SANU.</p></li>
</ul></td>
<td><p><code>p = 4.005e-13</code></p>
<ul>
<li><p>4.005e-13 &lt; 0.05 <span class="math inline">→</span> slope is statistically significant. Data provide sufficient evidence to determine that there is a linear relationship between PANU and EENU.</p></li>
</ul></td>
</tr>
<tr>
<td><p><span class="math inline"><em>R</em><sup>2</sup></span> = <code>0.009016</code></p>
<ul>
<li><p>Regression showed a very weak relationship between SANU and PANU. </p></li>
<li><p>SANU explains 0.9% of the variation in PANU. </p></li>
<li><p><span class="math inline"><em>R</em><sup>2</sup></span> = 1 indicates a perfect linear fit between the variables.</p></li>
</ul></td>
<td><p><span class="math inline"><em>R</em><sup>2</sup></span> = <code>0.225</code></p>
<ul>
<li><p>Regressions shows a moderate relationship between EENU and PANU, stronger than SANU and PANU.</p></li>
<li><p>EENU explains 22.5% of the variation in PANU. </p></li>
</ul></td>
</tr>
</tbody>
</table>

### c). Pearson's r correlation test: PANU-SANU and PANU-EENU

```{r}
# 1. Pearson correlation between PANU and SANU

cor.test(nonuse_clean$PANU,nonuse_clean$SANU, method = "pearson")

# 2. Pearson correlation between PANU and EENU

cor.test(nonuse_clean$PANU,nonuse_clean$EENU, method = "pearson")
```

<table>
<thead>
<tr>
<th><p>PANU-SANU</p></th>
<th><p>PANU-EENU</p></th>
</tr>
</thead>
<tbody>
<tr>
<td><p><code>r = 0.09495356</code></p>
<ul>
<li><p>r is very close to 0, indicating a very weak positive linear association between SANU and PANU.</p></li>
</ul></td>
<td><p><code>r = 0.4743524</code></p>
<ul>
<li><p>r indicates a moderate positive linear correlation between EENU and PANU.</p></li>
</ul></td>
</tr>
</tbody>
</table>

### d). Result Interpretation

PANU was not significantly associated with SANU ($r = 0.095$), ($p = 0.171$). The linear regression model showed a very weak positive relationship, with SANU explaining only around 0.9% of the variance in PANU ($R^2 = 0.009$). On the otherhand, PANU was significantly associated with EENU and the linear regression model showed a moderate positive association with EENU ($r = 0.474$), ($p < 0.001$). The PANU–EENU regression model explained around 22.5% of the variance in PANU ($R^2 = 0.225$). These results suggest that elbow extension non-use (EENU) is more strongly associated with proximal arm non-use (PANU) than shoulder antepulsion non-use (SANU) in this dataset.\

Several limitations should be considered including:

- Pearson's correlation and linear regression only assess linear relationships, so a possible non-linear association may not be detected.

- The scatterplots contain several observations far from the main cluster, and these potential outliers may influence both the correlation coefficients and regression models.

- The EENU model explains only 22.5% of the variance in PANU, indicating that EENU alone does not fully explain PANU. 77.5% is still left unexplained.

- Correlation and regression demonstrate association, not causation, so these results cannot show that EENU causes changes in PANU. People with greater EENU tend to also have greater PANU, which is an association, but the model can not say that a greater EENU causes greater PANU.

  - **Association:** two variables tend to change together.

  <!-- -->

  - **Causation:** changing one variable actually produces a change in the other.

- Rows containing missing values were removed from the analysis, which could introduce bias if the missing data were not random.

## 6. **Statistical tests:** what are statistical tests?

<table>
<colgroup>
<col style="width: 17%" />
<col style="width: 61%" />
<col style="width: 21%" />
</colgroup>
<thead>
<tr>
<th><p>Test</p></th>
<th><p>Definition</p></th>
<th><p>Reference</p></th>
</tr>
</thead>
<tbody>
<tr>
<td><p><strong>Mean</strong></p></td>
<td><p>The sum of all values divided by the total number of values. </p></td>
<td><p>Ch. 1, 1.4.1 <em>Measures of center: mean and median</em>, p. <strong>30</strong></p></td>
</tr>
<tr>
<td><p><strong>Median</strong> </p></td>
<td><p>The middle value when the observations are ordered from smallest to largest </p></td>
<td><p>Ch. 1, 1.4.1 <em>Measures of center: mean and median</em>, p. <strong>30</strong></p></td>
</tr>
<tr>
<td><p><strong>Variance</strong> </p></td>
<td><p>A measure of how spread out the values are around the mean. A large variance means the values are more spread out.</p></td>
<td><p>Ch. 1, 1.4.2 <em>Measures of spread</em>, p. <strong>31</strong></p></td>
</tr>
<tr>
<td><p><strong>Standard deviation</strong></p></td>
<td><p>Square root of the variance. Another measure of spread. How far values tend to be from the mean, in the same units as the data.</p></td>
<td><p>Ch. 1, 1.4.2 <em>Measures of spread</em>, p. <strong>31</strong></p></td>
</tr>
<tr>
<td><p><strong>Normal distribution</strong> </p></td>
<td><p>A symmetric, bell-shaped distribution where most values are close to the mean. A perfectly normal distribution, the mean, median, and mode are equal.</p></td>
<td><p>Ch. 3, 3.3 <em>Normal distribution</em>, pp. <strong>152–153</strong></p></td>
</tr>
<tr>
<td><p><strong>Statistical test</strong></p></td>
<td><p>A method used to decide whether a difference or relationship in the sample is likely to be real or could just be due to random chance. usually starting with a null hypothesis   (<span class="math inline"><em>H</em><sub>0</sub></span>). </p>
<p>used when we want to determine whether an observed difference/association in a sample provides sufficient evidence of a corresponding effect in the population, while accounting for sampling variability. </p></td>
<td><p>Ch. 4, 4.3 <em>Hypothesis testing</em>, p. <strong>212-214</strong></p></td>
</tr>
<tr>
<td><p><strong>Parametric test</strong></p></td>
<td><p>A statistical test used for normally distributed data and makes more assumptions about the data.</p>
<p>Assumptions: independent observations, a normal distribution and for some group comparisons, similar variances between groups. </p></td>
<td><p>Ch. 4, 4.3 <em>Hypothesis testing</em>, p. <strong>212-214</strong></p>
<p>Ch. 5, pp. <strong>238, 245, 248</strong></p></td>
</tr>
<tr>
<td><p><strong>Non-parametric test</strong></p></td>
<td><p>A statistical test used for non-normally distributed data and makes fewer assumptions. </p>
<p>Assumptions: Not normally distributed, appropriately measured data, independence of observations when comparing independent groups.</p></td>
<td><p>Ch. 4, 4.3 <em>Hypothesis testing</em>, p. <strong>212-214</strong></p>
<p>Ch. 5, pp. <strong>238, 245, 248</strong></p></td>
</tr>
<tr>
<td><p> <strong>p-value</strong></p></td>
<td><p>Tells us how unusual our result would be if there were actually no real effect or difference. A small p-value means the result would be unlikely under the null hypothesis.</p></td>
<td><p>Ch. 4, 4.3.1, Step 4, p. <strong>214</strong></p></td>
</tr>
<tr>
<td><p><strong>Paired test</strong></p></td>
<td><p>Used when observations are related, such as measurements taken from the same participants before and after treatment.</p></td>
<td><p>Ch. 5, 5.2 <em>Two-sample test for paired data</em>, pp. <strong>244–246</strong></p></td>
</tr>
<tr>
<td><p><strong>Unpaired test</strong> </p></td>
<td><p>Used when the groups are independent, containing different observations or participants.</p></td>
<td><p>Ch. 5, 5.3 <em>Two-sample test for independent data</em>, pp. <strong>247–250</strong></p></td>
</tr>
</tbody>
</table>

**Risk of error when using a statistical test? (**Ch. 4, 4.3.4 *Decision errors*, pp. **220–221)**

- **Type I error** = You say there is an effect when there actually is not.

  - when we reject a true null hypothesis (false positive), with probability ($\alpha$).

- **Type II error** = You say there is no effect when there actually is one.

  - when we fail to reject a false null hypothesis (false negative), with probability ($\beta$)

**Reference**

Vu, J., & Harrington, D. (2020). *Introductory statistics for the life and biomedical sciences* (1st ed.). OpenIntro. [biostat statistics textbook](https://1drv.ms/b/c/e002ca8c8522705b/IQBV-YZ48EdOR4CDjGYRw5cAASO4MgXNelHfJSpdsDT3GpI?e=gp5aNh)

## 7. **Effect of treatment over time**

Q: Does a treatment training have an effect over time?

Treatment is a rehabilitation training for individuals with cardiac conditions.

### a). Import data

```{r}

# 1.Import the data
  
  prepost <- read_csv("data/PrePost.csv")
  
  # View the first 6 rows to make sure it was imported correctly 
  head(prepost)

```

### b). Check the structure of the data

8 measurements before rehabilitation and 8 after:

```         
Before: 125 130 132 135 136 138 140 145
After:  127 132 133 136 139 141 145 148
```

```{r}

# Display the variable names
names(prepost)

# Check the structure and data types
str(prepost)

# Obtain a first summary of the values
summary(prepost)
```

### c). Variables of the dataset

| Variables | Description | Unit | Possible Values | Expected Change |
|----------|-----------------|--------|------------------|--------------------|
| perf | Performance measurement taken before and after cardiac rehabilitation | Not specified | Numerical values; observed in this dataset from 125 to 148 | Expected to increase if a higher performance score represents better physical capacity |
| time | Indicates whether the performance measurement was taken before or after rehabilitation | None | `Before`, `After` | Not applicable; this is the measurement occasion |

### d). Data analysis

Q: Does the rehabilitation treatment have an effect on measured performance from before to after treatment?

**a) What comparison are we making?**

- performance before rehabilitation vs performance after rehabilitation

  - Before: 125 130 132 135 136 138 140 145
  - After: 127 132 133 136 139 141 145 148

- If the same 8 participants are being compared before and after then this is a paired comparison

**b) Parametric or non-parametric test?**

- For a paired test, what matters is the distribution of the differences:
  - $difference = After - Before$
  - The differences are \[2 2 1 1 3 3 5 3\]
- Check that the differences are normally distributed using a `Shapiro-Wilk normality test`.
  - If $p > 0.05$ = not statistically significant and there is not enough evidence to say the differences depart from normality (we do not reject normality) $\rightarrow$ **`parametric test`** $\rightarrow$ `paired t-test`
  - If $p < 0.05$ = statistically signifcant and there is enough evidence to say the differences are not normally distributed (we reject normality) $\rightarrow$ **`non-parametric test`** $\rightarrow$ **`Wilcoxon signed-rank test`**.
  - For this dataset $p = 0.283 > 0.05$ therefore the data is normally distributed and a parametric paired t-test can be used for statistical analysis since the data are paired.
- The `paired t-test` gives $p = 0.001 < 0.05$ $\rightarrow$ there is a statistically significant difference between performance before and after rehabilitation.

```{r}
# 1. Calculate the differences between before and after in the dataset

# a) Separate the before and after performance values
before <- prepost %>% filter(time == "Before") %>% pull(perf) 
# only keep rows where time == "Before" and extract the perf column 

after <- prepost %>% filter(time == "After") %>% pull(perf)
# only keep rows where time == "After" and extract the perf column 

# b) Calculate the difference for each participant
difference <- after - before

difference

```

```{r}
# 2. Check for normality of data using Shapiro-Wilk normality test

shapiro.test(difference)

```

```{r}
# 3. Paired t-test comparing performance before and after rehabilitation
t.test(after, before, paired = TRUE)
```

**c) Which graph illustrates the effect of rehabilitation treatment?**

Since this is paired data, use a paired line plot to best show the effect of rehabilitation treatment on each participant.

```{r}

# 1. Reorganize data structure so each participant has a id (number 1-8)

# a). Add a new column to dataset: participant IDs to label each participant 
prepost_id <- prepost %>% mutate(id = rep(1:8, 2))  # give participants a # 1-8 and repeat twice
 

# 2. Create the paired line plot
ggplot(prepost_id, aes(x = time, y = perf, group = id, color = factor(id))) + 
  #color = factor(id) → converts numbers to categorical data to label each person with a colour
  # group = id → connect the correct person's points
  
  geom_line() +
  geom_point() +
  labs(title = "Performance Before and After Rehabilitation", 
       x = "Time", y = "Performance", color = "Participant") + # create legend labeled participant
  theme_minimal()


```

**d) Rehabilitation treatment interpretation**

Mean performance increased after rehabilitation compared with before. The slightly larger standard deviation after rehabilitation indicates that participant performance values were slightly more spread out around the mean.

- **Before rehabilitation:** `135.1 ± 6.2` = mean ± standard deviation

- **After rehabilitation:** `137.6 ± 7.0` = mean ± standard deviation

  - Smaller SD = participants values are more similar to each other

  - Larger SD = participant values are more spread out.

```{r}

# 1. Calculate the mean & SD before rehabilitation

mean(before)
sd(before)

# 2. Calculate the mean & SD after rehabilitation

mean(after)
sd(after)

```

## 8. **Testing Stereotypes**

### a). Import data

```{r}

# 1.Import the data

snore <- read_delim("data/snore.csv") # delim seperates the columns correctly 

# View the first 6 rows to make sure it was imported correctly 
head(snore)
```

### b). Data analysis

Q: Do the data confirm the following stereotypes? 

**a) Are snorers fatter?**

The BMI of snorers and non-snores can be compared to gives us a measure of body mass relative to height, so it is more appropriate than simply asking whether snorers weigh more.

$BMI_{snorers}$ vs $BMI_{non-snoreres}$

$BMI = \frac{weight(kg)}{height (m)^2}$

```{r}

# 1. Create a new column BMI with the height m 
snore <- snore %>% mutate(BMI = weight / (height / 100)^2) # mutate adds a new column BMI 

head(snore)

```

```{r}
# 2. Make a boxplot to see the distribution of data before deciding the type of statistical test

ggplot(snore, aes(x = snore, y = BMI)) +
  geom_boxplot() +
  labs(title = "BMI of Snorers and Non-Snorers", x = "Snoring", y = "BMI (kg/m²)") +
  theme_minimal()
```

```{r}

# 3. Create a table to look ath the descriptive statistics
snore %>% group_by(snore) %>%
  summarise(mean_BMI = mean(BMI), median_BMI = median(BMI), sd_BMI = sd(BMI), n = n())
```

Both groups are extremely similar and the boxplot also shows a very similar distribution and spread in both groups. Therefore descriptively, there is no obvious evidence that snorers have a higher BMI.

- Non-snorers (`N`): mean BMI = 27.27, median = 28.07, SD = 2.81, (n=65)

- Snorers (`O`): mean BMI = 27.11, median = 28.08, SD = 2.83, (n=35)

These are **unpaired/independent groups**, because the snorers and non-snorers are different people.

Before deciding the type of statistical test: check to see if BMI is normally distributed in each group using the Shapiro-Wilk test.

```{r}
# 4. Check the normality of BMI in both groups using Shapiro-Wilk test.

# a) Check normality of BMI for non-snorers
shapiro.test(snore$BMI[snore$snore == "N"]) # take the BMI column only for the rows where snore = no 

# b) Check normality of BMI for snorers
shapiro.test(snore$BMI[snore$snore == "O"]) # take the BMI column only for the rows where snore = yes

```

BMI is not normally distributed in either group. Both are much smaller than (0.05), so we reject the assumption of normality for both groups:

$BMI_{snoreres}$ $\rightarrow$ $p = 2.323e-08 < 0.05$

$BMI_{non-snoreres}$ $\rightarrow$ $p = 2.323e9.81-07 < 0.05$

Since the two groups are independent and the data are non-normal, use the `Wilcoxon rank-sum test` (also called the `Mann–Whitney U test`). This tests whether the BMI distributions differ between snorers and non-snorers.

```{r}

# 5. Run the Wilcoxon rank-sum test on BMI 

wilcox.test(BMI ~ snore, data = snore)
```

$p = 0.951 > 0.05$ $\rightarrow$ there is no statistically significant difference in BMI between snorers and non-snorers. The BMI distributions of snorers and non-snorers are not meaningfully different in this sample.

> BMI was similar between snorers and non-snorers, with median values of 28.08 and 28.07 kg/m², respectively. The difference was not statistically significant (Wilcoxon rank-sum test, W = 1128.5, p = 0.951), providing no evidence to support the stereotype that snorers are fatter.

**b) Do snorers drink or smoke more?**

1.  **Do snorers drink more?** Alcohol consumption of snorers and non-snorers can be compared to determine whether snorers have higher alcohol consumption. Analyze the distribution of alcohol consumption before statistical analyses

    $Alcohol_{snorers}$ vs $Alcohol_{non-snorers}$

```{r}

# 1. Make a boxplot to compare alcohol consumption between snorers and non-snorers
ggplot(snore, aes(x = snore, y = alcool)) +
  geom_boxplot() +
  labs(title = "Alcohol Consumption of Snorers and Non-Snorers", x = "Snoring", y = "Alcohol") +
  theme_minimal()
```

```{r}
# 2. Create a table to look at the descriptive statistics

snore %>%
  group_by(snore) %>%
  summarise(mean_alcool = mean(alcool), median_alcool = median(alcool), sd_alcool = sd(alcool),   n = n())
```

> Snorers appear to consume more alcohol than non-snorers. Snorers had a higher mean alcohol consumption (4.03 vs 2.37) and a higher median (4 vs 0). The boxplot also shows generally higher alcohol values among snorers.

- Non-snorers (`N`): mean alcohol = 2.37, median = 0, SD = 3.18

- Snorers (`O`): mean alcohol = 4.03, median = 4, SD = 3.47

```{r}
# 3. Check the normality of alcohol consumption in both groups using the Shapiro-Wilk test

# a) Check normality for non-snorers
shapiro.test(snore$alcool[snore$snore == "N"])

# b) Check normality for snorers
shapiro.test(snore$alcool[snore$snore == "O"])
```

> Alcohol consumption was not normally distributed in either snorers or non-snorers (Shapiro–Wilk tests, both p \< 0.05). Therefore, a non-parametric test was used to compare the two unpaired/independent groups.

$Alcohol_{snoreres}$ $\rightarrow$ $p = 6.818e-09 < 0.05$

$Alcohol_{non-snoreres}$ $\rightarrow$ $p = 0.005032 < 0.05$

```{r}
# 4. Run the Wilcoxon rank-sum test on alcohol consumption

wilcox.test(alcool ~ snore, data = snore)
```

This result is statistically significant, meaning there is a difference between the two groups.

$p = 0.008594 < 0.05$

Snorers had higher alcohol consumption than non-snorers, with median values of 4 and 0, respectively. This difference was statistically significant (Wilcoxon rank-sum test, W = 788, p = 0.009), supporting the stereotype that snorers drink more.

2.  **Do snorers smoke more?**

Alcohol consumption of smokers and non-smokers can be compared to determine whether smokers have higher alcohol consumption. Analyze the proportion of smokers between snorers and non-snorers and since smoking and alcohol is categorical (`O`/`N`) data, therefore a histogram was plotted rather than a box-pot.

$Smoker_{non-snorer}$ vs $Smoker_{snorer}$

```{r}
# 1. Compare smoking distribution between snorers and non-snorers

ggplot(snore, aes(x = snore, fill = tabac)) +
  geom_bar(position = "fill") +
  labs(title = "Smoking Status of Snorers and Non-Snorers", x = "Snoring", y = "Proportion",      fill = "Smoking") +
  theme_minimal()
```

```{r}
# 2. Create a contingency table

# a). Calculate proportions within each snoring group

prop.table(smoke_table, margin = 1) # Calculate the percentages within each row, each row adds up to 100% 

# b). Convert the proportions to %

prop.table(smoke_table, margin = 1) * 100
```

> The descriptive statistics suggest that snorers do not smoke more in this sample. About 57.1% of snorers were smokers compared with 67.7% of non-snorers, so the proportion of smokers was actually lower among snorers.

- Among non-snorers, 32.31% are non-smokers and 67.69% are smokers

- Among snorers, 42.86% are non-smokers and 57.14% are smokers

Because both variables are categorical data (`O/N`), a `Chi-square test of independence` instead of a parametric or non-parametric test since the data is categorical and both groups are independent:

```{r}
# 3. Test whether smoking status is associated with snoring status

chisq.test(smoke_table)

```

No statistically significant association between smoking status and snoring status. Descriptively, snorers actually smoked less often in this sample (57.1% vs 67.7%)., but the Chi-square test shows that this difference is not statistically significant ($\chi^2(1)=0.689, p=0.407$), so the data do not support the stereotype that snorers smoke more.

$p=0.407>0.05$

**c) Are men fatter?**

We can compare BMI between men and women. The distribution of BMI between men and women are first analysed using a box-plot before statistical analyses is performed.

$BMI_{men}$ vs $BMI_{women}$

```{r}
# 1. Make a boxplot to compare BMI between men and women

ggplot(snore, aes(x = sex, y = BMI)) +
  geom_boxplot() +
  labs(title = "BMI of Men and Women", x = "Sex", y = "BMI (kg/m²)") +
  theme_minimal()
```

```{r}
# 2. Create a table to look at the descriptive statistics

snore %>%
  group_by(sex) %>%
  summarise(mean_BMI = mean(BMI), median_BMI = median(BMI), sd_BMI = sd(BMI), n = n())
```

Descriptive results suggest that men have a slightly higher BMI than women, but the difference looks small. The median BMI was 28.08 kg/m² for men and 27.73 kg/m² for women, while the mean BMI was 27.27 and 27.04 kg/m², respectively.

- Women (`F`): mean BMI = **27.04**, median = **27.73**, SD = **2.72**, (n=25)

- Men (`H`): mean BMI = **27.27**, median = **28.08**, SD = **2.85**, (n=75)

> Check the normality of the data using the Shapiro-Wilk test

```{r}

# 3. Check normality of BMI in both groups using the Shapiro-Wilk test

# a) Women
shapiro.test(snore$BMI[snore$sex == "F"])

# b) Men
shapiro.test(snore$BMI[snore$sex == "H"])
```

> BMI was not normally distributed in either women or men (Shapiro–Wilk tests, both p \< 0.05). Therefore, a non-parametric test was used to compare BMI between the two independent groups.

Men: $p = 1.154e-05 < 0.05$

Women: $p = 3.456e-09 < 0.05$

```{r}

# 4. Run the Wilcoxon rank-sum test on BMI

wilcox.test(BMI ~ sex, data = snore)
```

> No statistically significant difference in BMI between men and women.

$p = 0.488 > 0.05$

> Men had a slightly higher median BMI than women (28.08 vs 27.73 kg/m²), but the difference was not statistically significant (Wilcoxon rank-sum test, W = 850, p = 0.488). Therefore, the data do not support the stereotype that men are fatter.

**d) Do women smoke less?**

Both gender and `smoking` are categorical variables therefore the proportion of smokers among women and men must be compared.

$Smokers_{men}$ vs $Smokers_{women}$

```{r}

# 1. Make a proportional bar plot to compare smoking status between women and men

ggplot(snore, aes(x = sex, fill = tabac)) +
  geom_bar(position = "fill") +
  labs(title = "Smoking Status of Women and Men", x = "Sex", y = "Proportion", 
  fill = "Smoking") +
  theme_minimal()
```

```{r}
# 2. Create a contingency table for sex and smoking status

# a) Calculate proportions within each sex

prop.table(sex_smoke_table, margin = 1)

# b) Convert the proportions to percentages

prop.table(sex_smoke_table, margin = 1) * 100
```

A lower proportion of women were smokers compared with men (40% vs 72%). Descriptively, this supports the stereotype that women smoke less.

- **Women (`F`)**: 40% smokers, 60% non-smokers

- **Men (`H`)**: 72% smokers, 28% non-smokers

Since both gender and smoking are categorical variables a Chi-square test is performed

```{r}
# 3. Test whether smoking status is associated with sex

chisq.test(sex_smoke_table)

```

There is significant association between sex and smoking status.

$p=0.008<0.05$

A lower proportion of women were smokers compared with men (40% vs 72%). This difference was statistically significant ($\chi^2(1)=7.00, p=0.008$), supporting the stereotype that women smoke less in this sample.

**e) From a more general perspective: are there any correlations between variables?**
