---
fontsize: 9pt
format:
  typst:
    papersize: a4
    margin:
      x: 1.2cm
      y: 1.2cm
---

## A. Clinical Problem

**RQ:** Is proximal arm non-use (**PANU**) related to shoulder non-use (**SANU**) and or elbow non-use (**EENU**) after stroke?

### Variables

After a stroke, a patient may compensate for reduced shoulder or elbow use by leaning the trunk forward more. Therefore, we want to know whether:

- more **shoulder non-use** is associated with more **proximal arm non-use**
- more **elbow non-use** is associated with more **proximal arm non-use**

This can be studied using the **correlations between PANU, SANU and EENU**.

| Variable | Description | Unit | Range |
|----------|---------------------------------------|---------|------------------|
| SANU | shoulder flexion the patient can use with the trunk restrained but does not use spontaneously. | ° | -180° to 180° |
| EENU | elbow extension the patient can use with the trunk restrained but does not use spontaneously. | ° | -180° to 180° |
| PANU | shoulder and elbow movement together that the patient can use with the trunk restrained but does not use spontaneously. | \% | -100 to 100 |

**+** = non-use : patient used more movement in the restrained condition than in the spontaneous condition

**-** = over-use : patient used more movement spontaneously than in the restrained condition

**0** → no difference between the two conditions.

PANU (%) = (hand displacement − trunk displacement) / hand displacement

## B. Experimental Methodology

Participants with a previous stroke performed a **seated reaching task with the affected arm** while their movement was recorded using motion capture.

### Condition 1: Spontaneous

- The participant reaches for the cone using their natural movement strategy.
- Trunk compensation is allowed.
- This shows **how the participant normally chooses to perform the movement**.

### Condition 2: Trunk Restrained

- The participant performs the same reach while keeping the trunk still.
- This reduces forward trunk compensation.
- The participant must rely more on **shoulder and elbow movement**.
- This shows **what the affected arm can do when trunk compensation is restricted**.

## C. Data: How Non-Use Is Identified

- If the participant produces more shoulder and or elbow movement when the trunk is restrained than during spontaneous reaching this shows that they were **capable of using that movement but did not use it spontaneously**.
- **Non-use =** movement capacity available − movement used spontaneously

| Variable | Correlation Prediciton | Why? | Unit | Range |
|---------|---------------|-------------------------------------|-------|---------|
| SANU | Positive | More shoulder non-use should mean more overall proximal-arm non-use. | \% | 0-100 |
| EENU | Positive | More elbow non-use should mean more overall proximal-arm non-use. | \% | 0-100 |
| PANU | Positive | Both contribute to reaching, but one joint can partly compensate for the other. | \% | 0-100 |

# Data mining and statistical tests

## 1. Import data

```{r # Import libraries and import the data}
  
# Load package
library(tidyverse) # tools for plotting and data analysis

# Import the dataset
nonuse <- read_csv("data/NonUse.csv")

# View the first 6 rows to make sure it was imported correctly 
head(nonuse)
```

```{r}
# summary of the data
summary(nonuse)
```

## 2. Check the imported data

```{r}
str(nonuse) # structure of data
```

## 3. Clean data

The data set contains missing values, only complete pairs of observations were used for each analysis. Thus the data must be cleaned before performing any correlation analysis.

```{r}

# Keep only complete observations for all three variables
nonuse_clean <- nonuse %>%
  select(PANU, SANU, EENU) %>%
  drop_na() # removes rows with NA
```

```{r}
head(nonuse_clean) # check the new data table
```

## 4. Distribution of data

Before the correlation analysis, boxplots were used to inspect the distribution, spread, and possible outliers of PANU, SANU, and EENU.

```{r}

# 1. Convert the data from wide to long format
nonuse_long <- nonuse_clean %>%
  pivot_longer(cols = c(PANU, SANU, EENU), names_to = "Variable", values_to = "Value")
  # convert the data into a two columns one with variable names and the other with values because ggplot requires this data formating 

# 2. Create one boxplot for all three variables
ggplot(nonuse_long, aes(x = Variable, y = Value)) + 
  geom_boxplot() +
  labs(title = "Distribution of PANU, SANU and EENU", x = "Variable", y = "Value") +
  theme_minimal()

```

## 5. Correlation analysis

- Visualize `PANU–SANU` and `PANU–EENU`

- Calculate the correlations

- Interpret the correlation coefficients

- Perform the two linear regressions

- Extract the regression equations

- Write a scientific-results sentence

- Discuss limitations

### a). Visualize relationships: scatterplot of PANU-SANU and PANU-EENU

Q: As SANU or EENU changes, does PANU tend to change as well?

- `SANU or EENU` as the **x-variable**

- `PANU` as the **y-variable**

```{r}
# 1. Create a scatterplot for PANU-SANU

ggplot(nonuse_clean, aes(x = SANU, y = PANU)) +
  geom_point() + # represent each observation as a point
  labs(title = "Relationship between PANU and SANU", x = "SANU", y = "PANU") +
  theme_minimal()
```

```{r}
# 2. Create a scatterplot for PANU-EENU

ggplot(nonuse_clean, aes(x = EENU, y = PANU)) +
  geom_point() + 
  labs(title = "Relationship between PANU and EENU", x = "EENU", y = "PANU") +
  theme_minimal()
```

### Interpretation

> The PANU–EENU relationship appears more clearly positive, while the PANU–SANU relationship shows greater dispersion. Several observations are far from the main cluster, which may influence the correlation. The strength of these relationships will be quantified using correlation coefficients.

### b). Linear regression: PANU-SANU and PANU-EENU

General form: `y = ax + b`

***Metrics:***

1.  $p$ \< 0.05 $\rightarrow$ statistically significant
2.  $R^2$ = proportion of variation explained by the linear model
    - $R^2$ $\rightarrow$ (0 to 1)
3.  $r$ = direction + strength of linear association
    - $r$ $\rightarrow$ (-1 to +1)

```{r}
# 1. Linear regression: PANU-SANU

model_SANU <- lm(PANU ~ SANU, data = nonuse_clean) # store the linear model lm(). 
# PANU = left side of equation (variable we want to predict), SANU = right side of equation (predictor variable)

summary(model_SANU) # show the main regression results (intercept b, slope a)

# 2. Linear regression: PANU-EENU

model_SANU <- lm(PANU ~ EENU, data = nonuse_clean)

summary(model_SANU) 
```

<table>
<thead>
<tr>
<th><p>PANU-SANU</p></th>
<th><p>PANU-EENU</p></th>
</tr>
</thead>
<tbody>
<tr>
<td><p>PANU = a * SANU + b </p></td>
<td><p>PANU = a * EENU + b </p></td>
</tr>
<tr>
<td><p><code>PANU = 0.10093 * SANU + 8.86819</code></p>
<ul>
<li><p>Slope (a) = 0.10093 <span class="math inline">→</span>  (a &gt;0) positive relationship </p></li>
<li><p>Intercept (b) = 8.86819</p></li>
</ul></td>
<td><p><code>PANU = 0.39851 * EENU + 5.67295</code></p>
<ul>
<li><p>Slope (a) = 0.39851 <span class="math inline">→</span>  (a &gt;0) positive relationship </p></li>
<li><p>Intercept (b) = 5.67295</p></li>
</ul></td>
</tr>
<tr>
<td><p><code>p = 0.171</code> </p>
<ul>
<li><p>0.171 &gt; 0.05 <span class="math inline">→</span> slope is not statistically significant. Although the slop is positive, the data do not provide sufficient evidence to determine a linear relationship between PANU and SANU.</p></li>
</ul></td>
<td><p><code>p = 4.005e-13</code></p>
<ul>
<li><p>4.005e-13 &lt; 0.05 <span class="math inline">→</span> slope is statistically significant. Data provide sufficient evidence to determine that there is a linear relationship between PANU and EENU.</p></li>
</ul></td>
</tr>
<tr>
<td><p><span class="math inline"><em>R</em><sup>2</sup></span> = <code>0.009016</code></p>
<ul>
<li><p>Regression showed a very weak relationship between SANU and PANU. </p></li>
<li><p>SANU explains 0.9% of the variation in PANU. </p></li>
<li><p><span class="math inline"><em>R</em><sup>2</sup></span> = 1 indicates a perfect linear fit between the variables.</p></li>
</ul></td>
<td><p><span class="math inline"><em>R</em><sup>2</sup></span> = <code>0.225</code></p>
<ul>
<li><p>Regressions shows a moderate relationship between EENU and PANU, stronger than SANU and PANU.</p></li>
<li><p>EENU explains 22.5% of the variation in PANU. </p></li>
</ul></td>
</tr>
</tbody>
</table>

### c). Pearson's r correlation test: PANU-SANU and PANU-EENU

```{r}
# 1. Pearson correlation between PANU and SANU

cor.test(nonuse_clean$PANU,nonuse_clean$SANU, method = "pearson")

# 2. Pearson correlation between PANU and EENU

cor.test(nonuse_clean$PANU,nonuse_clean$EENU, method = "pearson")
```

<table>
<thead>
<tr>
<th><p>PANU-SANU</p></th>
<th><p>PANU-EENU</p></th>
</tr>
</thead>
<tbody>
<tr>
<td><p><code>r = 0.09495356</code></p>
<ul>
<li><p>r is very close to 0, indicating a very weak positive linear association between SANU and PANU.</p></li>
</ul></td>
<td><p><code>r = 0.4743524</code></p>
<ul>
<li><p>r indicates a moderate positive linear correlation between EENU and PANU.</p></li>
</ul></td>
</tr>
</tbody>
</table>

### d). Result Interpretation

PANU was not significantly associated with SANU ($r = 0.095$), ($p = 0.171$). The linear regression model showed a very weak positive relationship, with SANU explaining only around 0.9% of the variance in PANU ($R^2 = 0.009$). On the otherhand, PANU was significantly associated with EENU and the linear regression model showed a moderate positive association with EENU ($r = 0.474$), ($p < 0.001$). The PANU–EENU regression model explained around 22.5% of the variance in PANU ($R^2 = 0.225$). These results suggest that elbow extension non-use (EENU) is more strongly associated with proximal arm non-use (PANU) than shoulder antepulsion non-use (SANU) in this dataset.\

Several limitations should be considered including:

- Pearson's correlation and linear regression only assess linear relationships, so a possible non-linear association may not be detected.

- The scatterplots contain several observations far from the main cluster, and these potential outliers may influence both the correlation coefficients and regression models.

- The EENU model explains only 22.5% of the variance in PANU, indicating that EENU alone does not fully explain PANU. 77.5% is still left unexplained.

- Correlation and regression demonstrate association, not causation, so these results cannot show that EENU causes changes in PANU. People with greater EENU tend to also have greater PANU, which is an association, but the model can not say that a greater EENU causes greater PANU.

  - **Association:** two variables tend to change together.

  <!-- -->

  - **Causation:** changing one variable actually produces a change in the other.

- Rows containing missing values were removed from the analysis, which could introduce bias if the missing data were not random.
