---
fontsize: 9pt
format:
  typst:
    papersize: a4
    margin:
      x: 1.2cm
      y: 1.2cm
---

## A. Clinical Problem

**RQ:** Is proximal arm non-use (**PANU**) related to shoulder non-use (**SANU**) and or elbow non-use (**EENU**) after stroke?

### Variables

After a stroke, a patient may compensate for reduced shoulder or elbow use by leaning the trunk forward more. Therefore, we want to know whether:

- more **shoulder non-use** is associated with more **proximal arm non-use**
- more **elbow non-use** is associated with more **proximal arm non-use**

This can be studied using the **correlations between PANU, SANU and EENU**.

| Variable | Description | Unit | Range |
|----------|---------------------------------------|---------|------------------|
| SANU | shoulder flexion the patient can use with the trunk restrained but does not use spontaneously. | ° | -180° to 180° |
| EENU | elbow extension the patient can use with the trunk restrained but does not use spontaneously. | ° | -180° to 180° |
| PANU | shoulder and elbow movement together that the patient can use with the trunk restrained but does not use spontaneously. | \% | -100 to 100 |

**+** = non-use : patient used more movement in the restrained condition than in the spontaneous condition

**-** = over-use : patient used more movement spontaneously than in the restrained condition

**0** → no difference between the two conditions.

PANU (%) = (hand displacement − trunk displacement) / hand displacement

## B. Experimental Methodology

Participants with a previous stroke performed a **seated reaching task with the affected arm** while their movement was recorded using motion capture.

### Condition 1: Spontaneous

- The participant reaches for the cone using their natural movement strategy.
- Trunk compensation is allowed.
- This shows **how the participant normally chooses to perform the movement**.

### Condition 2: Trunk Restrained

- The participant performs the same reach while keeping the trunk still.
- This reduces forward trunk compensation.
- The participant must rely more on **shoulder and elbow movement**.
- This shows **what the affected arm can do when trunk compensation is restricted**.

## C. Data: How Non-Use Is Identified

- If the participant produces more shoulder and or elbow movement when the trunk is restrained than during spontaneous reaching this shows that they were **capable of using that movement but did not use it spontaneously**.
- **Non-use =** movement capacity available − movement used spontaneously

| Variable | Correlation Prediciton | Why? | Unit | Range |
|---------|---------------|-------------------------------------|-------|---------|
| SANU | Positive | More shoulder non-use should mean more overall proximal-arm non-use. | \% | 0-100 |
| EENU | Positive | More elbow non-use should mean more overall proximal-arm non-use. | \% | 0-100 |
| PANU | Positive | Both contribute to reaching, but one joint can partly compensate for the other. | \% | 0-100 |

# Data mining and statistical tests

## 1. Load data

```{r # Import libraries and import the data}
  
# Load package
library(tidyverse) # tools for plotting and data analysis

# Import the dataset
nonuse <- read_csv("data/NonUse.csv")

# View the first 6 rows to make sure it was imported correctly 
head(nonuse)
```

```{r}
# summary of the data
summary(nonuse)
```

## 2. Check the imported data

```{r}
str(nonuse) # structure of data
```

## 3. Clean data

The data set contains missing values, only complete pairs of observations were used for each analysis. Thus the data must be cleaned before performing any correlation analysis.

```{r}

# Keep only complete observations for all three variables
nonuse_clean <- nonuse %>%
  select(PANU, SANU, EENU) %>%
  drop_na() # removes rows with NA
```

```{r}
head(nonuse_clean) # check the new data table
```

## 
